---
title: "gov51-ps3"
author: "Angela Calderon Pio"
format: pdf
---

<https://github.com/angelacalderonpio/gov51-ps3.git>

# 1.1

```{r}

#1.1

library(tidyverse)
library(glmnet)
library(knitr)
library(ggplot2)
library(dplyr)
library(tinytex)

compas <- read.csv("data/raw/compas_clean.csv")

dim(compas)


```

# 1.2 

There are 7,214 observations and 17 variables are in dataset. The unit of observation is an individual person (in this case, a person/defendant in the COMPAS dataset).

# 1.3

```{r}

mean(compas$recidivism, na.rm = TRUE)

```

a\) The overall recidivism rate in the data is 0.4506515.

b\) This outcome is binary, which matters for interpretation because a linear model assumes a continuous outcome. Consequently, predicted values can be less than 0 or greater than 1 which are not meaningful probabilities.

# 1.4

```{r}

means <- sapply(compas, function(x) mean(as.numeric(x), na.rm = TRUE))
sds <- sapply(compas, function(x) sd(as.numeric(x), na.rm = TRUE))

summary_table <- data.frame(
  variable = names(compas),
  mean = means,
  sd = sds
)

summary_table

```

The variables age and priors_count have the most variation at standard deviations of 11.88892178 and 4.88253835, respectively. The variables with the least variation are native_american and asian with standard deviations of 0.04989256 and 0.06645867, respectively.

# 1.5

```{r}

compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count %in% 1:2 ~ "1-2",
     priors_count %in% 3:5 ~ "3-5",
     priors_count %in% 6:10 ~ "6-10",
     priors_count >= 11 ~ "11+"
  ),
  priors_cat = factor(priors_cat, levels = c("0", 
                                             "1-2", "3-5", "6-10", "11+"))) |>
  group_by(priors_cat) |>
  summarize( rate = mean(recidivism, na.rm = TRUE)) |>
  ggplot(aes(x = priors_cat, y = rate)) +
  geom_col(color = "black", fill = "pink") +
  labs(
    x = "Number of Prior Offenses",
    y = "Recidivism Rate"
  )

ggsave("output/q1.5_plot.png", width = 6, height = 4)


```

The pattern shows that as the number of prior offenses increases, the recidivism rate also increases, suggesting a positive relationship between these two variables.

\newpage

# 2.1

```{r}

set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor (0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]
nrow(train)
nrow(test)

```

In the train set, there are 5,771 observations. In the test set, there are 1,443 observations.

# 2.2

```{r}

model <- lm(recidivism ~ ., data = train)
summary(model)


```

The predictors that are statistically significant at the 5% level include age, female, priors_counts, felony, has_juv_misd, and age_25_45, as their p-values are below 0.05.

Each additional prior offense is associated with about a 0.03 increase in the predicted probability of recidivism, holding all other variables constant.

# 2.3

```{r}


sqrt(1/nrow(train) * sum((predict(model, newdata = train) - train$recidivism)^2))

sqrt(1/nrow(test) * sum((predict(model, newdata = test) - test$recidivism)^2))

0.4605811 - 0.462227

```

The gap between the in-sample and the out-of-sample RMSE is -0.0016459. This is very small, suggesting that the model performs similarly on both the training and test data, so OLS with 16 predictors does not appear to be overfitting.

# 2.4

```{r}

set.seed(51)

folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)
for (k in 1:5) {
  train_k <- train[folds != k, ]
  test_k <- train[folds == k, ]
  model_k <- lm(recidivism ~ ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)

0.4620144- 0.462227
```

a\) The 5-fold cross-validation RMSE is 0.4620144. This is quite similar to the single-split out-of-sample RMSE (0.462227), indicating that the model's performance is stable across different samples. (There is a 0.0002126 difference).

b\) Cross-validation is more reliable as it utilizes the mean across multiple splits rather than relying on one random split.

c\) Cross-validation is useful for choosing $\lambda$ because it tests different values on new data and picks the one with the lowest prediction error.

\newpage

# 3.1

```{r}

# Full interaction formula
f_full <- recidivism ~ (.)^2 +
I(age^2) + I(priors_count^2) +
I(juv_fel_count^2) + I(juv_misd_count^2) +
I(juv_other_count^2)
# Create design matrices for glmnet (we'll need these later)
X_train <- model.matrix(f_full, data = train)[, -1]
X_test <- model.matrix(f_full, data = test)[, -1]
y_train <- train$recidivism
y_test <- test$recidivism
# How many features now?
ncol(X_train)

5771/141
5771/16

```

a\) The expanded model has 141 variables.

b\) The ratio of training observations to predictors is about 40.92908:1 In the simple model, the ratio was higher at 360.6875:1. Thus, the expanded model uses more predictors relative to the sample size.

# 3.2

```{r}

ols_full <- lm(f_full, data = train)

sqrt(1/5771 * sum((predict(ols_full, newdata = train) - train$recidivism)^2))

sqrt(1/1443 * sum((predict(ols_full, newdata = test) - test$recidivism)^2))

0.4485474-0.4890411

```

a\) In sample RMSE: 0.4485474

b\) Out-of-sample RMSE: 0.4890411

c\) The gap between in-sample and out-of-sample RMSE is larger than in the simple model (here, 0.0404937. The in-sample RMSE is lower (0.44854740, but the out-of-sample RMSE is higher (0.4890411), indicating worse performance on new data. This suggests overfitting, as the expanded model is too complex and fits noise in the training data as opposed to general patterns.

# 3.3

This phenomenon is called overfitting. Adding many interaction terms makes the model itself more flexible, so it can fit the training data more closely and reduce in-sample error, But, it begins to capture random noise rather than actual patterns, leading to worse performance on new data (or out-of sample).

\newpage

# 4.1

```{r}

set.seed(51) # for reproducible CV folds
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
cv_ridge

plot(cv_ridge)

ggsave("output/q4.1_plot.png", width = 6, height = 4)


```

a\) The optimal $\lambda$ selected by cross-validation is 0.0138

b\) The plot shows how prediction error changes with $\lambda$. Error decreases at first and then flattens, and the optimal $\lambda$ is the one that gives the lowest error.

# 4.2

```{r}

pred_ridge <- predict(cv_ridge, s = "lambda.min", newx = X_test)

sqrt(mean((y_test - pred_ridge)^2))


```

The out-of-sample RMSE for Ridge is 0.4700385. This is lower than the kitchen-sink OLS (which was 0.4890411) but slightly higher than the simple OLS (around 0.4620144). This suggests that Ridge reduces overfitting compared to the expanded model through shrinking coefficients, but does not outperform the simpler model here.

# 4.3

```{r}

coef(cv_ridge, s = "lambda.min")

```

No, Ridge did not eliminate any variables. While some coefs are very small, none are exactly zero. Conceptually this is sound as Ridge shrinks coefficients toward zero, but doesn't make them equal zero.

\newpage

# 5.1

```{r}

set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)

cv_lasso

plot(cv_lasso)

ggsave("output/q5.1_plot.png", width = 6, height = 4)

```

a\) The optimal $\lambda$ is 0.007694.

b\) The plot shows that the error initially falls, then reaches its minimum around $\lambda$ = 0.007694, then begins to rise again.

# 5.2

```{r}

coef_lasso <- coef(cv_lasso, s = "lambda.min")
coef_lasso

sum(coef_lasso != 0)
sum(coef_lasso == 0)

```

a\) Out of the 141 variables, 28 survived: thus, 114 were eliminated.

b\) A few of the surviving interaction terms include:

-   female:black

-   black:juv_other_count

-   priors_count:age_under_25

-   other_race:has_juv_felony

These make substantive sense because they capture the effect of how criminal history or demographics could potentially vary across different groups.

c\) This illustrates Baker's argument because LASSO automatically selects which variables to keep based on predictive performance, rather than relying on a person to choose them manually (which might yield a degree of subjectivity). By shrinking coefficients to zero, it reduces subjective decisions about which interactions matter.

# 5.3

```{r}

pred_lasso <- predict(cv_lasso, s = "lambda.min", newx = X_test)

sqrt(mean((y_test - pred_lasso)^2))

```

The LASSO out-of-sample RMSE is 0.4592597. This is lower than both Ridge (0.4700385) and the kitchen-sink OLS (0.4890411), and slightly better than the simple OLS (0.4620144). This indicates that LASSO reduces overfitting and improves prediction.

\newpage

# 6.1

```{r}

set.seed(51)

cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)

cv_enet

pred_enet <- predict(cv_enet, s = "lambda.min", newx = X_test)
sqrt(mean((y_test - pred_enet)^2))

coef_enet <- coef(cv_enet, s = "lambda.min")
sum(coef_enet != 0)

```

The out of sample RMSE for the Elastic Net is 0.45383446, which is slightly lower than LASSO, Ridge, and both OLS models. It keeps more variables than LASSO but far fewer than Ridge, balancing variable selection and shrinkage. It retains 43 variables.

# 6.2

```{r}

results <- data.frame(
  Model = c("OLS (main effects, 16 vars)",
            "OLS (kitchen sink , ~ 141 vars)", "Ridge", "LASSO", "Elastic Net"),
  Predictors = c(16, 141, 141, 28, 43),
  RMSE = c(0.462, 0.489, 0.470, 0.459, 0.458)
)

kable(results)

```

a\) The model with the lowest out-of-sample RMSE is the Elastic Net model, with a value of 0.458. This indicates the best predictive performance.

b\) The kitchen-sink OLS model has the best in-sample fit because it uses the most variables so it can match the training data very closely (and unlike Ridge, it doesn't shrink coefficients toward zero). However, it is not the sample model with the best out-of-sample performance as it fits to the noise in the data.

c\) This table demonstrates bias-variance tradeoff: the kitchen-sink OLS model has low bias but high variance, leading to overfitting and sub-ideal out-of-sample performance. The other models like Ridge, LASSO, and Elastic Net reduce variance by shrinking or choosing coefficients, but increases bias. Elastic Net seems to be the best model here as it balances both bias and variance.

# 6.3

a\) Lasso helped narrow the disagreement by removing the need for experts to subjectively choose peer firms. That is, instead of manually picking comparison companies, LASSO uses data to select which firms actually improve predictive accuracy. Because it makes the process more objective, experts are more likely to wind up with similar models and conclusions.

b\) Based on this problem set, the biggest source of manipulation in traditional OLS is the choice of which variables to include in the model. In this approach, you can include many predictors even if they are weak or irrelevant, which can make the model fit the training data superbly, but hurt out-of-sample performance. Because OLS doesn't restrict these choices, analysts can impact results through this.

# 6.4

a\) No, the LASSO coefficients on priors_count would not answer the causal question. It is designed to improve prediction, not identify causal effects, and it might shrink/removal variables based on predictive usefulness rather than causal importance. The estimate could further be biased by omitted variables, so it can't tell us whether priors_count itself causes higher recidivism.

b\) The difference between using regression for prediction versus causal inference is that prediction focuses accuracy whereas causal inference focuses on identifying actual cause and effect relationships. More specifically, the former seeks to forecast outcomes, even if coefs are biased. But in the latter, research design is important so that we can actually interpret the coefs as causal effects.

# 6.5

```{r}

test$pred <- predict(model, newdata = test)

mean(test$pred[test$black ==1])

mean(test$pred[
  test$black == 0 &
    test$hispanic == 0 &
    test$asian == 0 &
    test$other_race == 0 &
    test$native_american == 0
])

0.511985 - 0.3789606

sum(test$black == 1)
sum(
  test$black == 0 &
  test$hispanic == 0 &
  test$asian == 0 &
  test$other_race == 0 &
  test$native_american == 0
)

```

a\) Yes, there is a difference. For Black defendants, the mean predicted recidivism probability is 0.511985, whereas for white defendants, it is 0.3789606 (the difference being 0.1330244).

```{r}

mean(test$recidivism[test$black ==1])

mean(test$recidivism[
  test$black == 0 &
    test$hispanic == 0 &
    test$asian == 0 &
    test$other_race == 0 &
    test$native_american == 0
])

0.521262 - 0.4020202

```

The actual recidivism rates for Black and white defendants are 0.521262 and 0.4020202, respectively. Compared to the predicted rates, the model assigns a higher predicted probability to Black defendants than to white defendants, so the predictions reflect a similar pattern to the actual data.

c\) There are pros and cons to using predictive algorithms in criminal justice. One on hand, they can improve consistency and help identify individuals at higher risk of recidivism. On the other hand, as seen in this question, both the predicted and actual recidivism rates were higher for Black defendants, which raises concerns that these models may reflect or reinforce existing disparities in data. And I think generally, it is important to recognize that these are imperfect predictive models. The goal is to ensure fairness and improve predictive accuracy.
