---
title: "gov51-ps3"
author: "Angela Calderon Pio"
---

<https://github.com/angelacalderonpio/gov51-ps3.git>

```{r}

#1.1

library(tidyverse)
library(glmnet)
library(knitr)
library(ggplot2)
library(dplyr)

compas <- read.csv("data/raw/compas_clean.csv")

dim(compas)


```

# 1.2 

There are 7,214 observations and 17 variables are in dataset. The unit of observation is an individual person (in this case, a person/defendant in the COMPAS dataset).

# 1.3

```{r}

mean(compas$recidivism, na.rm = TRUE)

```

a\) The overall recidivism rate in the data is 0.451.

b\) This outcome is binary, which matters for interpretation because a linear model assumes a continuous outcome. Consequently, predicted values can be less than 0 or greater than 1 which are not meaningful probabilities.

# 1.4

```{r}

means <- sapply(compas, function(x) mean(as.numeric(x), na.rm = TRUE))
sds <- sapply(compas, function(x) sd(as.numeric(x), na.rm = TRUE))

summary_table <- data.frame(
  variable = names(compas),
  mean = means,
  sd = sds
)

summary_table

```

The variables age and priors_count have the most variation at standard deviations of 11.90 and 4.883, respectively. The variables with the least variation are native_american and asian with standard deviations of 0.050 and 0.066, respectively.

# 1.5

```{r}

compas |>
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count %in% 1:2 ~ "1-2",
     priors_count %in% 3:5 ~ "3-5",
     priors_count %in% 6:10 ~ "6-10",
     priors_count >= 11 ~ "11+"
  ),
  priors_cat = factor(priors_cat, levels = c("0", "1-2", "3-5", "6-10", "11+"))) |>
  group_by(priors_cat) |>
  summarize( rate = mean(recidivism, na.rm = TRUE)) |>
  ggplot(aes(x = priors_cat, y = rate)) +
  geom_col(color = "black", fill = "pink") +
  labs(
    x = "Number of Prior Offenses",
    y = "Recidivism Rate"
  )


```

The pattern shows that as the number of prior offenses increases, the recidivism rate also increases, suggesting a positive relationship between these two variables.
